\documentclass[11pt]{article}

% --- Layout and typography ------------------------------------------------
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{microtype}                 % better justification, far fewer overfull lines
\usepackage[nobottomtitles*]{titlesec} % never strand a section heading at the foot of a page
\renewcommand{\bottomtitlespace}{0.16\textheight}
\usepackage{amsmath,amssymb,bm}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}

% Prefer whitespace at the foot of a page to stretched or broken paragraphs.
\raggedbottom
\widowpenalty=10000
\clubpenalty=10000
\displaywidowpenalty=10000
\setlist[itemize]{itemsep=3pt,topsep=4pt,parsep=0pt}

% --- Notation -------------------------------------------------------------
\newcommand{\R}{\mathbf{R}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bmm}{\mathbf{m}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\bK}{\mathbf{K}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bF}{\mathbf{F}}
\newcommand{\bP}{\mathbf{P}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bG}{\mathbf{G}}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bL}{\mathbf{L}}
\newcommand{\bnu}{\bm{\nu}}
\newcommand{\skewm}[1]{\lfloor #1 \rfloor_\times}
\newcommand{\dtheta}{\delta\bm{\theta}}
\newcommand{\dx}{\delta\bx}
\newcommand{\real}{\mathbb{R}}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\Log}{Log}
\DeclareMathOperator{\atantwo}{atan2}
\DeclareMathOperator{\Ex}{\mathbb{E}}

\title{\textbf{A Consistency-Gated Quaternion Error-State Kalman Filter\\[2pt]
for Multi-Sensor Inertial Navigation}}
\author{Irfan Saric}
\date{December 2025}

\begin{document}
\maketitle

\begin{abstract}
\noindent
We describe the design, derivation, implementation, and verification of an error-state Kalman filter
(ESKF) that fuses a strapdown inertial measurement unit (IMU) with up to nine aiding
modalities---GPS position and Doppler velocity, a barometric altimeter, a magnetometer, a downward
laser altimeter (LiDAR), ultra-wideband (UWB) radio ranging, optical flow, a Doppler velocity log
(DVL), and a direct attitude fix---to estimate the pose of an aerial vehicle. The filter follows the
multiplicative, local error-state formulation of Sol\`a~\cite{sola2017} and Trawny and
Roumeliotis~\cite{trawny2005}: a nominal state carries the full estimate and is integrated directly
from the IMU, while a minimal $15$-dimensional error state, reset to zero after every correction,
carries the covariance. We derive the continuous-time error dynamics from first principles, discretize
them, and give each of the nine measurement models with its exact error-state Jacobian; every Jacobian
is validated against a finite difference. The central claim is not merely that the filter tracks well
but that its \emph{reported uncertainty is correct}: we verify this with the normalized estimation
error squared (NEES) consistency test of Bar-Shalom et~al.~\cite{barshalom2001}, run as a Monte-Carlo
gate over many simulated flights and shown to hold throughout each flight, not only in the mean, and we
reject outliers online with the companion normalized innovation squared (NIS) test. On the full
nine-sensor fusion the filter is statistically consistent (position NEES $2.85$ against an expected
$3.0$; full-state NEES $16.2$ against $15.0$) at \SI{0.05}{\metre} position RMSE, while remaining
faithful to inertial-navigation error and observability theory: with all aiding removed the position
error grows without bound, and it is recovered to sub-metre accuracy by any single position-fixing
modality, including GPS-denied UWB ranging. A recorded reference dataset and a replay tool accompany
the filter, all reported figures are generated directly by the estimator, and the complete
system---core, simulator, and a real-time 3D visualization---runs in the browser via WebAssembly and
WebGL2.
\end{abstract}

\section{Introduction}

Aided inertial navigation estimates a vehicle's position, velocity, and orientation by integrating
high-rate inertial measurements and correcting the accumulated drift with lower-rate aiding sensors
\cite{farrell2008,groves2013,titterton2004}. Inertial integration alone is untenable over any useful
horizon: accelerometer and gyroscope errors are integrated once into velocity and attitude and twice
into position, so a bias of even a milli-$g$ becomes metres of position error within seconds. Aiding
sensors---GPS, a barometer, a camera, a radio ranging system---bound that drift, but only if the fusion
that combines them is correct in a specific and often overlooked sense: it must report an honest
uncertainty.

Orientation makes the estimation problem subtle. A unit quaternion has four components but only three
degrees of freedom, so a filter that estimates it directly must either enforce the unit-norm
constraint or carry a covariance for a quantity that is not free to move in four directions; either
choice is awkward and, done naively, inconsistent. The \emph{error-state} (equivalently indirect, or
multiplicative) Kalman filter resolves this by estimating a minimal three-parameter rotation error in
the tangent space of the current attitude estimate~\cite{markley2003,lefferts1982,sola2017}. The
nominal orientation absorbs the full, unconstrained rotation; the filter's covariance describes only a
small perturbation about it, which is genuinely a three-vector and whose linearization is accurate
because the perturbation is kept near zero by frequent resets.

The distinction this paper insists on is between an estimator that \emph{tracks} well and one that is
\emph{consistent}. A tracker minimizes error; a consistent filter additionally reports a covariance
that matches that error in distribution. The difference matters because a covariance is not decoration:
downstream consumers---a path planner deciding whether a gap is safely traversable, a controller
weighting a control action by confidence, a second estimator fusing this one's output---all trust the
reported uncertainty. An overconfident filter (covariance too small) invites those consumers to act on
noise as though it were signal; an underconfident one wastes information. A subtly wrong Jacobian or a
mis-scaled process-noise term frequently leaves the RMSE almost unchanged while corrupting the
covariance, so tracking error alone cannot detect it. The normalized estimation error squared (NEES)
can, and it is the gate around which this work is organized.

\paragraph{Contributions.}
The contribution is one of rigor and reproducibility rather than novelty of formulation:
\begin{itemize}
  \item a complete, self-contained derivation of the local error-state kinematics, their
        discretization, and the exact error-state Jacobian of each of \textbf{nine} measurement models,
        with every Jacobian checked against a central finite difference to $10^{-4}$;
  \item a discussion of the \textbf{observability} of the fused system that explains, rather than
        merely reports, why each sensor suite recovers the states it does;
  \item a verification methodology built on the \textbf{NEES consistency gate}, evaluated both pooled
        and as a function of time, complemented by an online \textbf{NIS} outlier test;
  \item full \textbf{reproducibility}: a recorded reference dataset, a replay harness, and figure data
        emitted directly by the filter, so every number and curve in this paper is real estimator
        output;
  \item an implementation whose numerical core is dependency-free and free of unsafe code, and which
        runs unchanged from a native test harness to a real-time browser visualization.
\end{itemize}
Source, the dataset, and an interactive demonstration are available at
\url{https://github.com/Zarathos94/ekf-sim} and \url{https://irfansaric.com/showcase/ekf-sim}.

\paragraph{Outline.}
Section~\ref{sec:notation} fixes notation and frame conventions. Section~\ref{sec:related} surveys
related work and public datasets. Section~\ref{sec:state} defines the nominal and error states.
Sections~\ref{sec:kin} and \ref{sec:disc} derive and discretize the error dynamics.
Section~\ref{sec:meas} gives the nine measurement models, and Section~\ref{sec:obs} analyzes their
observability. Section~\ref{sec:inject} covers injection, reset, and numerical care;
Section~\ref{sec:nees} the consistency methodology. Sections~\ref{sec:sim}--\ref{sec:results} describe
the simulation environment and present results, and
Sections~\ref{sec:impl}--\ref{sec:conclusion} cover implementation, limitations, and conclusions.

\section{Notation and Frame Conventions}
\label{sec:notation}

Two reference frames are used: a local-level \emph{world} frame $W$ (a fixed east-north-up tangent
frame, with gravity along $-\be_3$) and a \emph{body} frame $B$ fixed to the vehicle. Vectors are
columns in $\real^3$; $\be_1,\be_2,\be_3$ are the standard basis. The orientation of the body in the
world is a \emph{Hamilton} unit quaternion $\bq\in\mathbb{S}^3$ with real part first,
$\bq=(q_w,\,q_x,\,q_y,\,q_z)$, and $\R=\R(\bq)\in SO(3)$ is the corresponding rotation matrix mapping
\emph{body to world}, so a body-frame vector $\bu_B$ appears in the world as $\R\bu_B$ and a world
vector in the body as $\R^\top\bu_W$. We use the Hamilton convention throughout (as
in~\cite{sola2017}); it differs from the JPL convention of~\cite{trawny2005} by the handedness of the
quaternion product and a transpose in $\R(\bq)$, a distinction that is a frequent source of sign errors
and is spelled out in Appendix~\ref{app:quat}.

The skew-symmetric (cross-product) operator $\skewm{\cdot}:\real^3\to\mathfrak{so}(3)$ is
\begin{equation}
  \skewm{\ba} =
  \begin{bmatrix} 0 & -a_3 & a_2 \\ a_3 & 0 & -a_1 \\ -a_2 & a_1 & 0 \end{bmatrix},
  \qquad \skewm{\ba}\bv = \ba\times\bv .
\end{equation}
The maps $\Exp:\real^3\to SO(3)$ and its inverse $\Log$ are the group exponential and logarithm
(Appendix~\ref{app:so3}). For a manifold state we write the generalized addition and subtraction
$\boxplus$ and $\boxminus$: for orientation, $\bq\boxplus\dtheta = \bq\otimes\Exp(\tfrac12\dtheta)$ and
$\bq_2\boxminus\bq_1 = \Log(\bq_1^{-1}\otimes\bq_2)$; for vector states they reduce to ordinary
$+$ and $-$. Expectation is $\Ex[\cdot]$, and $\bx\sim\mathcal N(\bm\mu,\bm\Sigma)$ denotes a Gaussian.
Hats denote estimates and a leading $\delta$ denotes an error quantity.

\section{Related Work and Datasets}
\label{sec:related}

The indirect Kalman filter for attitude dates to spacecraft navigation; its canonical form is the
multiplicative EKF (MEKF) of Lefferts, Markley, and Shuster~\cite{lefferts1982}, surveyed by
Markley~\cite{markley2003}, Crassidis et~al.~\cite{crassidis2007}, and, in quaternion form, by Trawny
and Roumeliotis~\cite{trawny2005} and Sol\`a~\cite{sola2017}; it is the attitude workhorse of aided-INS
texts~\cite{farrell2008,groves2013,titterton2004} and of the modern tutorial treatment of Kok et
al.~\cite{kok2017}. For visual-inertial estimation the multi-state constraint Kalman filter (MSCKF) of
Mourikis and Roumeliotis~\cite{mourikis2007} keeps a sliding window of poses, and optimization-based
systems such as VINS-Mono~\cite{qin2018} and the factor-graph / incremental-smoothing approach of
iSAM2 and preintegration~\cite{kaess2012,forster2017} trade recursion for re-linearization. The
invariant EKF of Barrau and Bonnabel~\cite{barrau2017} improves consistency by exploiting the group
structure of the state; the filter here uses the more common local-error parameterization but
demonstrates consistency empirically and diagnoses it through observability. Radio ranging with UWB for
micro-aerial vehicles follows Mueller et~al.~\cite{mueller2015}, and metric optical flow follows
Honegger et~al.~\cite{honegger2013}.

Consistency and observability are tightly linked in this literature. Huang et~al.~\cite{huang2010} and
Hesch et~al.~\cite{hesch2014} showed that a standard EKF for visual-inertial navigation gains spurious
information along the four truly unobservable directions of the problem (global position and heading),
which manifests as an \emph{inconsistent} filter; observability-constrained variants repair it. Closed-form
observability analyses of inertial fusion are given by Martinelli~\cite{martinelli2012}. We revisit these
results in Section~\ref{sec:obs} to explain our scenario outcomes.

Public datasets are the standard proving ground for such estimators: EuRoC MAV~\cite{burri2016} and
TUM-VI~\cite{schubert2018} provide synchronized IMU, stereo imagery, and millimetre-accurate motion
capture ground truth; UZH-FPV~\cite{delmerico2019} targets aggressive first-person-view flight; and
KITTI~\cite{geiger2013} provides automotive IMU, GNSS, and LiDAR. Because ground truth on such data is
itself an estimate and the sensor suites differ from ours, we validate primarily against an
\emph{analytic} trajectory---where truth is exact and the NEES is therefore meaningful to the last
digit---and additionally ship a recorded, replayable dataset (Section~\ref{sec:sim}) so that our
specific results are reproducible bit-for-bit.

\section{State Representation}
\label{sec:state}

The \textbf{nominal state} collects position $\bp\in\real^3$, velocity $\bv\in\real^3$, orientation
$\bq\in\mathbb{S}^3$ (body to world), and the accelerometer and gyroscope biases
$\ba_b,\bomega_b\in\real^3$:
\begin{equation}
  \bx = (\bp,\ \bv,\ \bq,\ \ba_b,\ \bomega_b).
\end{equation}
It carries the full, unconstrained estimate and is integrated directly from the IMU. The
\textbf{error state} is the minimal $15$-vector
\begin{equation}
  \dx = (\delta\bp,\ \delta\bv,\ \dtheta,\ \delta\ba_b,\ \delta\bomega_b) \in \real^{15},
\end{equation}
in which the rotation error $\dtheta$ is defined \emph{locally} (in the body frame) through the
multiplicative relation
\begin{equation}
  \bq_{\text{true}} = \bq \otimes \Exp\!\left(\tfrac{1}{2}\dtheta\right),
  \label{eq:localerror}
\end{equation}
so that the corresponding rotation matrices satisfy $\R_{\text{true}} = \R\,\Exp(\dtheta)$. The error
state is reset to zero after every correction, so its $15\times15$ covariance
$\bP\in\real^{15\times15}$ stays well conditioned and its orientation block is an ordinary $3\times3$
matrix rather than a rank-deficient $4\times4$ one. This separation---an unconstrained nominal state
plus a small, minimal, frequently reset error state---is the entire point of the error-state
formulation: the extended Kalman filter's linearization is applied only to the small quantity, where it
is accurate.

Figure~\ref{fig:arch} shows the overall dataflow and Algorithm~\ref{alg:eskf} the per-cycle loop; the
remainder of the paper fills in each block.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    >={Stealth[length=2.4mm]},
    box/.style={draw, rounded corners, align=center, minimum height=12mm, minimum width=23mm,
                font=\small, inner sep=4pt},
    lbl/.style={font=\scriptsize}]
    \node[box] (imu) {IMU\\$\ba_m,\bomega_m$};
    \node[box, right=10mm of imu] (prop) {Propagate\\$\hat{\bx}^-,\ \bP^-$};
    \node[box, right=13mm of prop] (corr) {Correct\\$\bK,\ \dx$};
    \node[box, right=13mm of corr] (state) {State\\$\hat{\bx}^+,\ \bP^+$};
    \node[box, below=11mm of corr] (aid) {Aiding sensors\\[-1pt]
      {\scriptsize GPS, baro, mag, LiDAR,}\\[-2pt]{\scriptsize UWB, flow, DVL, att.}};
    \draw[->] (imu) -- (prop);
    \draw[->] (prop) -- node[lbl,above]{$\bF\bP\bF^{\!\top}\!+\bQ$} (corr);
    \draw[->] (corr) -- node[lbl,above]{inject $\dx$} (state);
    \draw[->] (aid) -- node[lbl,right=1pt]{$\bz,\ \mathbf{R}$} (corr);
    % feedback routed clearly above the boxes -- it crosses no node
    \draw[->] (state.north) -- ++(0,8mm)
      -| node[lbl,above,pos=0.25]{next IMU sample; reset $\dx\!\gets\!\mathbf{0}$} (prop.north);
  \end{tikzpicture}
  \caption{The error-state filter. The IMU drives the nominal state forward and inflates the
  error-state covariance ($\bP^-\!=\!\bF\bP\bF^{\!\top}\!+\bQ$); each aiding measurement produces a
  correction $\dx$ that is injected into the nominal state, after which the error state is reset to zero
  and the cycle repeats on the next IMU sample.}
  \label{fig:arch}
\end{figure}

\begin{algorithm}[t]
\caption{Error-state Kalman filter, one cycle}
\label{alg:eskf}
\begin{algorithmic}[1]
\State \textbf{Predict} (each IMU sample): integrate $\bp,\bv,\bq$ (Eq.~\ref{eq:nominal});
       set $\bP \gets \bF\bP\bF^{\!\top}+\bQ$ (Eqs.~\ref{eq:F},~\ref{eq:Q})
\ForAll{available measurements $\bz$ with model $h$, Jacobian $\bH$, noise $\mathbf{R}$}
  \State $\bm{y} \gets \bz-h(\hat{\bx})$;\quad $\bS \gets \bH\bP\bH^{\!\top}+\mathbf{R}$
         \Comment{innovation and its covariance}
  \State \textbf{if} $\bm{y}^{\!\top}\bS^{-1}\bm{y} > \gamma$ \textbf{then} reject (NIS gate,
         Section~\ref{sec:nees})
  \State $\bK \gets \bP\bH^{\!\top}\bS^{-1}$;\quad $\dx \gets \bK\,\bm{y}$
  \State \textbf{Inject}: $\hat{\bx} \gets \hat{\bx} \boxplus \dx$;\quad
         $\bP \gets \bG\bP\bG^{\!\top}$;\quad reset $\dx \gets \mathbf{0}$
  \State $\bP \gets (\I-\bK\bH)\bP(\I-\bK\bH)^{\!\top}+\bK\mathbf{R}\bK^{\!\top}$ \Comment{Joseph form}
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Continuous-Time Kinematics and Error Dynamics}
\label{sec:kin}

We first write the true kinematics, then linearize them about the nominal state to obtain the dynamics
of the error state, which the filter propagates.

\paragraph{Nominal kinematics.}
Let $\ba_m$ and $\bomega_m$ be the accelerometer and gyroscope readings and define the bias-corrected
specific force and body rate $\ba = \ba_m - \ba_b$, $\bomega = \bomega_m - \bomega_b$. With gravity
$\bg = -g\,\be_3$ ($g=\SI{9.80665}{\metre\per\second\squared}$), the continuous nominal kinematics are
\begin{equation}
  \dot{\bp} = \bv, \qquad
  \dot{\bv} = \R\ba + \bg, \qquad
  \dot{\bq} = \tfrac12\,\bq \otimes \bomega, \qquad
  \dot{\ba}_b = \mathbf{0}, \qquad
  \dot{\bomega}_b = \mathbf{0},
  \label{eq:nomkin}
\end{equation}
where $\bomega$ in the quaternion derivative is treated as a pure quaternion $(0,\bomega)$. The biases
are modelled as constant in the mean and driven only by noise (below).

\paragraph{Error kinematics.}
Write each true quantity as the nominal plus an error: $\bp_t=\bp+\delta\bp$, $\bv_t=\bv+\delta\bv$,
$\R_t=\R\,\Exp(\dtheta)\approx\R(\I+\skewm{\dtheta})$, $\ba_{b,t}=\ba_b+\delta\ba_b$,
$\bomega_{b,t}=\bomega_b+\delta\bomega_b$. The IMU readings carry additive white noise
$\bn_a,\bn_\omega$ and the biases random walks $\bn_{ab},\bn_{\omega b}$. Substituting into
\eqref{eq:nomkin} and retaining first-order terms yields the continuous error dynamics (derived in the
manner of~\cite{sola2017,trawny2005}):
\begin{align}
  \dot{\delta\bp} &= \delta\bv, \label{eq:errp}\\
  \dot{\delta\bv} &= -\R\skewm{\ba}\,\dtheta - \R\,\delta\ba_b - \R\,\bn_a, \label{eq:errv}\\
  \dot{\dtheta}   &= -\skewm{\bomega}\,\dtheta - \delta\bomega_b - \bn_\omega, \label{eq:errth}\\
  \dot{\delta\ba_b} &= \bn_{ab}, \qquad
  \dot{\delta\bomega_b} = \bn_{\omega b}. \label{eq:errb}
\end{align}
The two coupling terms are the substance of the model. In \eqref{eq:errv}, perturbing the rotation in
$\R_t\ba_t = \R(\I+\skewm{\dtheta})(\ba-\delta\ba_b-\bn_a)$ and dropping second-order products gives the
term $\R\skewm{\dtheta}\ba = -\R\skewm{\ba}\dtheta$ (using $\skewm{\bm u}\bm w = -\skewm{\bm w}\bm u$):
an attitude error tilts the sensed specific force and thus corrupts the acceleration estimate, which is
the dominant mechanism of inertial position drift. In \eqref{eq:errth}, the local-error convention
makes the orientation-error dynamics depend on the body rate through $-\skewm{\bomega}$; this is the
term whose \emph{exact} discrete counterpart, rather than a first-order truncation, we retain in
Section~\ref{sec:disc}. Collecting \eqref{eq:errp}--\eqref{eq:errb} as
$\dot{\dx} = \bA\,\dx + \bL\,\bn$ gives the continuous system and noise matrices
\begin{equation}
  \bA =
  \begin{bmatrix}
    \mathbf 0 & \I & \mathbf 0 & \mathbf 0 & \mathbf 0\\
    \mathbf 0 & \mathbf 0 & -\R\skewm{\ba} & -\R & \mathbf 0\\
    \mathbf 0 & \mathbf 0 & -\skewm{\bomega} & \mathbf 0 & -\I\\
    \mathbf 0 & \mathbf 0 & \mathbf 0 & \mathbf 0 & \mathbf 0\\
    \mathbf 0 & \mathbf 0 & \mathbf 0 & \mathbf 0 & \mathbf 0
  \end{bmatrix},
  \qquad
  \bL =
  \begin{bmatrix}
    \mathbf 0 & \mathbf 0 & \mathbf 0 & \mathbf 0\\
    -\R & \mathbf 0 & \mathbf 0 & \mathbf 0\\
    \mathbf 0 & -\I & \mathbf 0 & \mathbf 0\\
    \mathbf 0 & \mathbf 0 & \I & \mathbf 0\\
    \mathbf 0 & \mathbf 0 & \mathbf 0 & \I
  \end{bmatrix},
  \label{eq:Acont}
\end{equation}
with $\bn=(\bn_a,\bn_\omega,\bn_{ab},\bn_{\omega b})$ having power spectral density
$\bQ_c=\diag(\sigma_a^2\I,\ \sigma_g^2\I,\ \sigma_{ab}^2\I,\ \sigma_{\omega b}^2\I)$.

\section{Discretization}
\label{sec:disc}

Over one IMU interval $\Delta t$ (rate $\approx\SI{200}{\hertz}$) the nominal state is integrated by
\begin{equation}
  \bp \leftarrow \bp + \bv\,\Delta t + \tfrac{1}{2}(\R\ba + \bg)\,\Delta t^2, \quad
  \bv \leftarrow \bv + (\R\ba + \bg)\,\Delta t, \quad
  \bq \leftarrow \bq \otimes \Exp\!\left(\tfrac{1}{2}\bomega\,\Delta t\right),
  \label{eq:nominal}
\end{equation}
with the quaternion renormalized after each step. The error-state covariance is propagated by
$\bP \leftarrow \bF\bP\bF^{\!\top} + \bQ$, where $\bF=\Exp_{\!A}(\bA\,\Delta t)$ is the discrete state
transition. Evaluating the matrix exponential blockwise, the nilpotent couplings contribute only their
first-order terms while the orientation block is kept exact:
\begin{equation}
  \bF =
  \begin{bmatrix}
    \I & \I\,\Delta t & \mathbf{0} & \mathbf{0} & \mathbf{0} \\
    \mathbf{0} & \I & -\R\skewm{\ba}\,\Delta t & -\R\,\Delta t & \mathbf{0} \\
    \mathbf{0} & \mathbf{0} & \Exp(-\bomega\,\Delta t) & \mathbf{0} & -\I\,\Delta t \\
    \mathbf{0} & \mathbf{0} & \mathbf{0} & \I & \mathbf{0} \\
    \mathbf{0} & \mathbf{0} & \mathbf{0} & \mathbf{0} & \I
  \end{bmatrix}.
  \label{eq:F}
\end{equation}
Here $\Exp(-\bomega\,\Delta t)=\exp(-\skewm{\bomega}\Delta t)$ is the exact rotation integral
(Appendix~\ref{app:so3}) rather than $\I-\skewm{\bomega}\Delta t$; the difference is third order in
$\bomega\Delta t$ but accumulates at the high body rates of aggressive flight, and---being an
orientation term---it is exactly the kind of small error the consistency test detects.

The discrete process noise is $\bQ = \int_0^{\Delta t}\!\bF(\tau)\,\bL\bQ_c\bL^{\!\top}\bF(\tau)^{\!\top}\,d\tau$,
which for a step this short is accurately approximated by $\bL\bQ_c\bL^{\!\top}\Delta t$ evaluated with
the block structure of \eqref{eq:Acont}, giving the diagonal form
\begin{equation}
  \bQ = \diag\!\big(\mathbf{0},\ \sigma_a^2\,\Delta t^2\,\I,\ \sigma_g^2\,\Delta t^2\,\I,\
  \sigma_{ab}^2\,\Delta t\,\I,\ \sigma_{\omega b}^2\,\Delta t\,\I\big).
  \label{eq:Q}
\end{equation}
The scaling deserves emphasis, because getting it wrong is the archetypal invisible bug. A per-sample
accelerometer noise of standard deviation $\sigma_a$ perturbs the integrated velocity by
$\sigma_a\Delta t$ over the step and therefore injects a \emph{variance} $\sigma_a^2\Delta t^2$; the
same holds for the orientation term. This is also the only dimensionally consistent choice: $\sigma_a$
has units of acceleration spectral density, and only $\sigma_a^2\Delta t^2$ produces a velocity
variance. The bias random walks, by contrast, accumulate variance linearly and scale as $\Delta t$. A
$\bQ$ built with the wrong power of $\Delta t$ can leave the RMSE almost unchanged while making the
filter badly overconfident or conservative---a discrepancy that is glaring in the NEES of
Section~\ref{sec:nees}, and which this test caught during development.

\section{Measurement Updates}
\label{sec:meas}

For a measurement $\bz = h(\bx) + \bnu$, $\bnu\sim\mathcal{N}(\mathbf 0,\mathbf{R})$, the error-state
Jacobian $\bH = \partial h / \partial \dx$ is evaluated at the nominal state and applied through
Algorithm~\ref{alg:eskf}. Because the error state is zero at the linearization point, the Jacobians
take a common form: a measurement that depends on orientation does so through $\R^\top\bm u$ for some
world vector $\bm u$, and the local perturbation $\R_t^\top = (\I-\skewm{\dtheta})\R^\top$ gives
$\R_t^\top\bm u = \R^\top\bm u + \skewm{\R^\top\bm u}\,\dtheta$ (using $-\skewm{\dtheta}\bm w =
\skewm{\bm w}\dtheta$), hence
\begin{equation}
  \frac{\partial (\R^\top\bm u)}{\partial\dtheta}\bigg|_{\dtheta=\mathbf 0} = \skewm{\R^\top\bm u},
  \label{eq:rotjac}
\end{equation}
the pattern shared by the magnetometer, optical-flow, and DVL updates below. The nine models, with the
non-zero blocks of $\bH$, are as follows.

\paragraph{GPS position and Doppler velocity.}
Absolute position and world-frame velocity, $h=\bp$ and $h=\bv$, with
$\partial h/\partial\delta\bp = \I_3$ and $\partial h/\partial\delta\bv = \I_3$ respectively. These are
the only absolute references and, as Section~\ref{sec:obs} shows, the states that make the rest
observable.

\paragraph{Barometer.}
Altitude only, $h=\be_3^\top\bp=p_z$, with $\partial h/\partial\delta\bp=\be_3^\top$.

\paragraph{Magnetometer.}
The world magnetic field $\bmm$ expressed in the body frame, $h=\R^\top\bmm$; by \eqref{eq:rotjac},
$\partial h/\partial\dtheta = \skewm{\R^\top\bmm}$. Treating the full field vector as the measurement,
rather than collapsing it to a scalar heading, keeps the update linear in the error state and avoids the
singularities of an explicit yaw extraction.

\paragraph{LiDAR altimeter.}
A downward beam ranging to flat ground senses the slant range $h = p_z / R_{33}$, where
$R_{33}=\be_3^\top\R\,\be_3$ is the cosine between the body-down and world-down axes. Then
$\partial h/\partial p_z = 1/R_{33}$, and differentiating $R_{33}$ through $\R_t=\R\,\Exp(\dtheta)$
gives $\partial h/\partial\dtheta = -\frac{p_z}{R_{33}^2}\,[-R_{32},\ R_{31},\ 0]$, so a single beam
senses height \emph{and} tilt.

\paragraph{UWB radio ranging.}
Scalar range to each fixed beacon $\bb_k$, $h=\lVert\bp-\bb_k\rVert$, with
$\partial h/\partial\delta\bp = \hat{\bu}^\top$, the unit line of sight from beacon to vehicle. A few
beacons at surveyed positions make absolute position observable with no GPS at
all~\cite{mueller2015}.

\paragraph{Optical flow.}
The body-frame horizontal velocity, $h = \Pi\,\R^\top\bv$ with $\Pi=[\I_2\ \ \mathbf 0]$ selecting the
first two components; then $\partial h/\partial\delta\bv = \Pi\R^\top$ and, by \eqref{eq:rotjac},
$\partial h/\partial\dtheta = \Pi\skewm{\R^\top\bv}$~\cite{honegger2013}.

\paragraph{Doppler velocity log (DVL).}
The full body-frame velocity, $h=\R^\top\bv$, with $\partial h/\partial\delta\bv = \R^\top$ and
$\partial h/\partial\dtheta = \skewm{\R^\top\bv}$.

\paragraph{Attitude fix.}
A direct orientation measurement $\bz_q$ (star tracker or vision): the innovation is the minimal
rotation $\bm y = \Log(\hat{\bq}^{-1}\!\otimes\bz_q) = \bz_q\boxminus\hat{\bq}$ and
$\partial h/\partial\dtheta = \I_3$---the standard multiplicative attitude update.

\medskip
Every Jacobian is checked in the test suite against a central finite difference across all $15$ error
directions, agreeing to $10^{-4}$; the attitude and velocity updates are additionally verified to
converge from a seeded error, which fails if a sign is wrong. All updates use the Joseph-form covariance
step of Algorithm~\ref{alg:eskf}, which remains symmetric positive-definite under finite precision.
Table~\ref{tab:sensors} summarizes the strapdown IMU and nine aiding modalities and the default noise
levels used throughout.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lccccl}
    \toprule
    Modality & Symbol dim. & Rate (\si{\hertz}) & Noise $\sigma$ & Frame & States aided \\
    \midrule
    IMU (accel.)        & $3$ & $200$ & \SI{0.06}{\metre\per\second\squared} & body & propagation \\
    IMU (gyro.)         & $3$ & $200$ & \SI{0.004}{\radian\per\second}       & body & propagation \\
    GPS position        & $3$ & $5$   & \SI{0.8}{\metre}      & world & $\bp$ \\
    GPS Doppler vel.    & $3$ & $5$   & \SI{0.1}{\metre\per\second} & world & $\bv$ \\
    Barometer           & $1$ & $20$  & \SI{0.6}{\metre}      & world & $p_z$ \\
    Magnetometer        & $3$ & $40$  & $0.02$ (unit)         & body  & $\dtheta$ (yaw) \\
    LiDAR altimeter     & $1$ & $25$  & \SI{0.15}{\metre}     & body  & $p_z,\dtheta$ \\
    UWB ranging         & $4$ & $10$  & \SI{0.35}{\metre}     & world & $\bp$ \\
    Optical flow        & $2$ & $30$  & \SI{0.15}{\metre\per\second} & body & $\bv,\dtheta$ \\
    DVL / radar vel.    & $3$ & $20$  & \SI{0.05}{\metre\per\second} & body & $\bv,\dtheta$ \\
    Attitude fix        & $3$ & $20$  & \SI{0.01}{\radian}    & --    & $\dtheta$ \\
    \bottomrule
  \end{tabular}
  \caption{The strapdown IMU and nine aiding modalities, with default rates and noise standard
  deviations. UWB reports one range per beacon (four beacons); the magnetometer and DVL/flow aid
  orientation through the coupling of \eqref{eq:rotjac}.}
  \label{tab:sensors}
\end{table}

\section{Observability}
\label{sec:obs}

Which states a given sensor suite can actually pin down---its \emph{observability}---explains the
results of Section~\ref{sec:results} far better than the RMSE numbers alone, and it is where an
error-state filter can quietly become inconsistent if the analysis is ignored~\cite{huang2010,hesch2014}.

With an absolute position reference (GPS or UWB), position and velocity are directly observable, and the
attitude becomes observable through the inertial coupling of \eqref{eq:errv}: an orientation error tilts
the sensed specific force, so as the vehicle accelerates the resulting position/velocity residual
reveals the tilt. Roll and pitch are additionally anchored by the gravity direction that the
accelerometer senses whenever the specific force is not purely gravitational. Heading (yaw) is the
delicate one: under constant-velocity motion with only inertial and position aiding it is
\emph{unobservable}, and the filter must not manufacture confidence about it---precisely the failure
mode that observability-constrained EKFs were introduced to prevent~\cite{hesch2014,martinelli2012}. A
magnetometer removes the ambiguity by measuring the world field directly (Section~\ref{sec:meas}); so
does sustained maneuvering, which couples yaw into the observable horizontal channels.

The gyroscope bias is observable whenever orientation is, and the accelerometer bias is observable when
the vehicle rotates, since rotation modulates how a fixed body-frame bias projects into the world
acceleration. This is why the analytic trajectory used here is a \emph{banking} helix rather than a
straight climb: the continuous change of attitude excites every bias direction and renders the full
$15$-state vector observable, which is a precondition for the NEES over all $15$ dimensions to be
meaningful.

These facts map directly onto the scenarios of Table~\ref{tab:scenarios}. Removing all aiding leaves an
unobservable, freely drifting integrator (dead reckoning). Restoring \emph{any} absolute position
sensor---GPS, or UWB ranging to surveyed beacons---restores observability of position and, through the
inertial coupling, of the rest; this is why GPS-denied UWB ranging recovers sub-metre accuracy, and why
the indoor suite (UWB, LiDAR, optical flow) needs no GPS. The vision-aided suite adds a direct attitude
fix, which makes orientation observable outright and tightens the attitude RMSE by nearly an order of
magnitude.

\section{Injection, Reset, and Numerical Care}
\label{sec:inject}

Once a correction $\dx$ is computed, it is folded into the nominal state and the error is reset to zero:
\begin{equation}
  \bp \leftarrow \bp + \delta\bp,\quad
  \bv \leftarrow \bv + \delta\bv,\quad
  \bq \leftarrow \bq\otimes\Exp\!\left(\tfrac12\dtheta\right),\quad
  \ba_b \leftarrow \ba_b + \delta\ba_b,\quad
  \bomega_b \leftarrow \bomega_b + \delta\bomega_b .
\end{equation}
Because the reset moves the linearization point of the orientation error onto the new nominal
quaternion, the covariance must be transported by the reset Jacobian
\begin{equation}
  \bG = \diag\!\big(\I,\ \I,\ \I-\tfrac12\skewm{\dtheta},\ \I,\ \I\big),
  \qquad \bP \leftarrow \bG\,\bP\,\bG^{\!\top}.
\end{equation}
Retaining the $\I-\tfrac12\skewm{\dtheta}$ term---frequently dropped, since it is within a whisker of
the identity for small corrections---is one of the details the consistency test rewards. Three further
numerical safeguards matter in finite precision: the covariance update uses the Joseph form
$(\I-\bK\bH)\bP(\I-\bK\bH)^{\!\top}+\bK\mathbf{R}\bK^{\!\top}$, which is positive-semidefinite by
construction where the shorter $(\I-\bK\bH)\bP$ form is not; the covariance is symmetrized as
$\tfrac12(\bP+\bP^{\!\top})$ after each step to suppress asymmetry from round-off; and the quaternion is
renormalized after every integration and injection so it never drifts off the unit sphere.

\section{Consistency: NEES and NIS}
\label{sec:nees}

The decisive test is \emph{statistical consistency}~\cite{barshalom2001}. Define the normalized
estimation error squared
\begin{equation}
  \varepsilon = \bm{e}^{\top}\bP^{-1}\bm{e}, \qquad
  \bm{e} = \hat{\bx}\boxminus\bx_{\text{true}} \in \real^{15},
\end{equation}
where the error is expressed in the filter's own tangent space: position and velocity differences, the
body-frame logarithm $\Log(\hat{\bq}^{-1}\!\otimes\bq_{\text{true}})$ for orientation, and bias
differences. For a consistent filter whose error is zero-mean Gaussian with covariance $\bP$, the
quadratic form is chi-square distributed, so $\Ex[\varepsilon]=\dim$ ($3$ for position, $15$ for the
full state). Averaged over $N$ independent Monte-Carlo runs, $N\bar\varepsilon \sim \chi^2(N\cdot\dim)$,
giving the two-sided $95\%$ acceptance interval
\begin{equation}
  \bar\varepsilon \in
  \left[\ \frac{\chi^2_{0.025}(N\cdot\dim)}{N},\ \frac{\chi^2_{0.975}(N\cdot\dim)}{N}\ \right].
  \label{eq:neesband}
\end{equation}
Overconfidence lands the average above the band, conservatism below. Two methodological points make the
test honest. First, the independent unit is the \emph{flight}, not the sample: NEES values within a run
are time-correlated, so the band is set by the number of runs, and pooling steady-state samples only
sharpens the estimate inside that band. Second, the runs are seeded from a realistic initial error drawn
from the initial covariance, so the test scores the filter rather than an omniscient initialization, and
the NEES is reported per five-second window as well as pooled, because consistency must hold throughout
the flight and not merely on average. The $\chi^2$ quantiles are evaluated with the Wilson--Hilferty
approximation, which is accurate to a fraction of a percent at the degrees of freedom used here.

The same machinery, applied to the \emph{innovation} rather than the estimation error, gives an online
outlier gate. The normalized innovation squared $\bm y^{\!\top}\bS^{-1}\bm y$ is $\chi^2$ with dimension
equal to that of the measurement, so a measurement whose NIS exceeds the upper quantile
$\gamma=\chi^2_{0.99}(\dim\bz)$ is rejected before it corrupts the state (Algorithm~\ref{alg:eskf},
line~4). Unlike the NEES, the NIS needs no ground truth and runs live in the browser.

\section{Simulation Environment and Test Data}
\label{sec:sim}

Ground truth is an analytic banking helical trajectory---a horizontal orbit of radius \SI{40}{\metre}
and period \SI{30}{\second} with a superimposed vertical oscillation---so position, velocity, and
acceleration are available in closed form and are exact. Orientation follows a coordinated turn (nose
along the velocity, rolled into the turn by the bank angle that balances lateral acceleration against
gravity), and the body angular rate is obtained by central-differencing the quaternion path. As
Section~\ref{sec:obs} noted, the continual banking is deliberate: it excites every bias direction and
makes the full state observable.

From this truth a strapdown IMU and the nine aiding sensors are synthesized, each with configurable
noise, bias random walk, rate, and availability (Table~\ref{tab:sensors}). A \texttt{record} command
writes the raw IMU stream, every aiding measurement, and ground truth to CSV; a \texttt{replay} command
re-runs the filter on that CSV and re-derives the RMSE and NEES, so the dataset is an independently
verifiable artifact rather than a live-only result. A \texttt{plotdata} command emits the exact series
plotted in the next section. The filter is deliberately given a process noise that differs from the
simulator's true noise---a filter never knows the truth---and it must remain consistent anyway.

\section{Results}
\label{sec:results}

All numbers and curves in this section are produced by the same \texttt{eskf} core that runs in the
browser, via the command-line harness.

\paragraph{Consistency.}
Table~\ref{tab:nees} reports the Monte-Carlo consistency check over $40$ flights of \SI{30}{\second}
with all nine sensors active. Both the position and full-state NEES fall inside the band of
\eqref{eq:neesband}. Figure~\ref{fig:nees} plots the ensemble-mean NEES as a function of time against
the same band: it neither climbs (which would signal a filter growing overconfident, the classic
inconsistency of an unconstrained EKF~\cite{huang2010}) nor sags, but sits on the expected value
throughout. The time-averaged RMSE is \SI{0.049}{\metre} in position, \SI{0.021}{\metre\per\second} in
velocity, and \SI{0.11}{\degree} in attitude; replaying the recorded dataset independently reproduces
\SI{0.046}{\metre}.

\begin{table}[t]
  \centering
  \begin{tabular}{lccc}
    \toprule
    Quantity & NEES mean & Expected & $95\%$ band \\
    \midrule
    Position (3-DOF)    & $2.85$  & $3.0$  & $[2.29,\ 3.81]$ \\
    Full state (15-DOF) & $16.23$ & $15.0$ & $[13.35,\ 16.74]$ \\
    \bottomrule
  \end{tabular}
  \caption{NEES consistency over $40$ Monte-Carlo flights, nine sensors active. Both statistics lie
  inside the band---the reported uncertainty is neither over- nor under-confident.}
  \label{tab:nees}
\end{table}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      name=posax, width=0.48\textwidth, height=4.6cm,
      xlabel={time (s)}, ylabel={position NEES}, xmin=5, xmax=30, ymin=1.8, ymax=4.2,
      tick label style={font=\footnotesize}, label style={font=\footnotesize},
      title style={font=\footnotesize}, title={3-DOF, expected $3.0$}]
      \fill[gray!15] (axis cs:5,2.29) rectangle (axis cs:30,3.81);
      \addplot[black, dashed, thick] coordinates {(5,3.0) (30,3.0)};
      \addplot[blue, thick] table[col sep=comma, x=t, y=pos] {figdata/nees_time.csv};
    \end{axis}
    \begin{axis}[
      name=fullax, at={($(posax.east)+(1.1cm,0)$)}, anchor=west,
      width=0.48\textwidth, height=4.6cm,
      xlabel={time (s)}, ylabel={full-state NEES}, xmin=5, xmax=30, ymin=12.5, ymax=18,
      tick label style={font=\footnotesize}, label style={font=\footnotesize},
      title style={font=\footnotesize}, title={15-DOF, expected $15.0$}]
      \fill[gray!15] (axis cs:5,13.35) rectangle (axis cs:30,16.74);
      \addplot[black, dashed, thick] coordinates {(5,15.0) (30,15.0)};
      \addplot[red, thick] table[col sep=comma, x=t, y=full] {figdata/nees_time.csv};
    \end{axis}
  \end{tikzpicture}
  \caption{Ensemble-mean NEES over $40$ flights as a function of time (blue/red), the expected value
  (dashed), and the $95\%$ acceptance band (shaded). The mean tracks the expected dimension for the whole
  flight; the filter is consistent throughout, not merely on average.}
  \label{fig:nees}
\end{figure}

\paragraph{Error against reported uncertainty.}
Figure~\ref{fig:envelope} shows a single flight's position error against the $\pm3\sigma$ envelope drawn
from the filter's own covariance, for the horizontal ($x$) and vertical ($z$) axes. The envelope starts
wide, reflecting the seeded initial uncertainty, and contracts as aiding information accumulates; the
error stays within it. An envelope that the error escaped, or one far larger than the error, would both
be visible here at a glance---this is the per-flight, per-axis view behind the aggregate NEES.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      name=exax, width=0.48\textwidth, height=4.6cm,
      xlabel={time (s)}, ylabel={$x$ error (m)}, xmin=0, xmax=30, ymin=-1.1, ymax=1.1,
      tick label style={font=\footnotesize}, label style={font=\footnotesize},
      legend style={font=\tiny, at={(0.98,0.02)}, anchor=south east}, legend cell align=left]
      \addplot[gray, dashed] table[col sep=comma, x=t, y=s3x] {figdata/envelope.csv};
      \addplot[gray, dashed] table[col sep=comma, x=t, y expr=-\thisrow{s3x}] {figdata/envelope.csv};
      \addplot[blue, thick] table[col sep=comma, x=t, y=ex] {figdata/envelope.csv};
      \legend{$\pm3\sigma$,, error}
    \end{axis}
    \begin{axis}[
      name=ezax, at={($(exax.east)+(1.1cm,0)$)}, anchor=west,
      width=0.48\textwidth, height=4.6cm,
      xlabel={time (s)}, ylabel={$z$ error (m)}, xmin=0, xmax=30, ymin=-0.8, ymax=0.8,
      tick label style={font=\footnotesize}, label style={font=\footnotesize}]
      \addplot[gray, dashed] table[col sep=comma, x=t, y=s3z] {figdata/envelope.csv};
      \addplot[gray, dashed] table[col sep=comma, x=t, y expr=-\thisrow{s3z}] {figdata/envelope.csv};
      \addplot[blue, thick] table[col sep=comma, x=t, y=ez] {figdata/envelope.csv};
    \end{axis}
  \end{tikzpicture}
  \caption{One flight's position error (blue) against its own $\pm3\sigma$ envelope (dashed), for the
  horizontal $x$ axis (left) and the vertical $z$ axis (right). The envelope contracts as the filter
  converges and contains the error---consistency seen for a single run.}
  \label{fig:envelope}
\end{figure}

\paragraph{Behaviour across sensor suites.}
Table~\ref{tab:scenarios} shows steady-state RMSE as the sensor suite is changed, and
Figure~\ref{fig:drift} contrasts the two extremes over time. With \emph{all} aiding removed the filter
dead-reckons on the IMU alone and the position error reaches hundreds of metres over
\SI{30}{\second}---the quadratic error growth that accelerometer and gyroscope bias predict for an
unaided strapdown system~\cite{groves2013,titterton2004}, and exactly what observability theory
requires of an unobservable integrator. Any single position-fixing modality restores sub-metre accuracy:
GPS-denied UWB ranging recovers from \SI{16.6}{\metre} to \SI{0.16}{\metre}, and the indoor and
vision-aided suites---using no GPS at all---stay well under a metre, consistent with the observability
argument of Section~\ref{sec:obs}.

\begin{table}[t]
  \centering
  \begin{tabular}{lccc}
    \toprule
    Scenario & Position (m) & Velocity (m/s) & Attitude (\si{\degree}) \\
    \midrule
    Nominal (GPS + baro + mag)         & $0.47$      & $0.24$  & $0.99$ \\
    IMU only (dead reckoning)          & $295$       & $25.4$  & $8.32$ \\
    GPS dropout                        & $16.6$      & $1.87$  & $2.05$ \\
    GPS denied $+$ UWB ranging         & $0.16$      & $0.10$  & $0.61$ \\
    Indoor (UWB $+$ LiDAR $+$ flow)    & $0.08$      & $0.04$  & $0.23$ \\
    Vision-aided (attitude $+$ DVL)    & $0.14$      & $0.03$  & $0.12$ \\
    Full fusion (all $9$ sensors)      & $0.06$      & $0.02$  & $0.11$ \\
    Noisy IMU ($5\times$)              & $0.48$      & $0.26$  & $1.06$ \\
    Bias drift ($10\times$)            & $0.52$      & $0.33$  & $1.53$ \\
    \bottomrule
  \end{tabular}
  \caption{Steady-state RMSE by scenario (mean of $12$ flights of \SI{30}{\second} each). Dead reckoning
  diverges; any single absolute position fix, including UWB alone, recovers it.}
  \label{tab:scenarios}
\end{table}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
      width=0.62\textwidth, height=5.2cm, ymode=log,
      xlabel={time (s)}, ylabel={position error (m)}, xmin=0, xmax=30, ymin=0.01, ymax=1000,
      tick label style={font=\footnotesize}, label style={font=\footnotesize},
      legend style={font=\footnotesize, at={(0.02,0.98)}, anchor=north west}, legend cell align=left]
      \addplot[red, thick] table[col sep=comma, x=t, y=imu_only] {figdata/drift.csv};
      \addplot[blue, thick] table[col sep=comma, x=t, y=aided] {figdata/drift.csv};
      \legend{IMU only (dead reckoning), full fusion}
    \end{axis}
  \end{tikzpicture}
  \caption{Position error versus time (log scale). Unaided inertial integration diverges to hundreds of
  metres as bias is integrated twice into position; the full nine-sensor fusion holds a few centimetres.
  The gap is five orders of magnitude by \SI{30}{\second}.}
  \label{fig:drift}
\end{figure}

\section{Implementation and Reproducibility}
\label{sec:impl}

The system is a small Rust workspace. The numerical core (\texttt{eskf}) has no dependencies, forbids
unsafe code, and carries its own fixed-size linear algebra, quaternion algebra, simulator, and NEES
metric; it is exercised by a native test suite that includes the finite-difference Jacobian checks and
the exp/log round-trips of Appendix~\ref{app:so3}. A command-line harness (\texttt{eskf-cli}) is the
correctness gate: \texttt{check} runs the Monte-Carlo NEES test of Section~\ref{sec:nees},
\texttt{scenarios} produces Table~\ref{tab:scenarios}, \texttt{record}/\texttt{replay} manage the
reference dataset, and \texttt{plotdata} emits the series behind Figures~\ref{fig:nees}--\ref{fig:drift}.
The identical core is compiled to WebAssembly (\texttt{eskf-wasm}) and driven by a hand-written WebGL2
front end that renders the estimate, ground truth, and live covariance ellipsoid in real time, with
per-sensor toggles and noise controls. Because the browser and the gate share one implementation, the
consistency proven at the command line is the consistency running in the demonstration. The reference
dataset and the figure data are committed alongside the source, so every table and figure in this paper
can be regenerated from a clean checkout with a single command.

\section{Limitations and Future Work}
\label{sec:limits}

The filter is a first-order (extended) estimator with additive Gaussian noise, and it inherits the usual
limitations. Linearization error grows with the update interval and the motion aggressiveness; an
iterated update or a sigma-point (unscented) propagation~\cite{julier2004} would reduce it at some cost.
The local-error parameterization is consistent here but is known to accrue spurious information along
unobservable directions in the harder visual-inertial case; the invariant/equivariant
EKF~\cite{barrau2017} addresses this at the level of the state's group structure and is the natural next
step. Measurements are assumed time-synchronized and in-sequence; real systems need delayed-state or
out-of-sequence handling. Finally, validation here is against a simulator with exact ground truth, which
is what makes the NEES meaningful; the natural complement is a run against a public benchmark with motion
capture, such as EuRoC~\cite{burri2016} or TUM-VI~\cite{schubert2018}, where the sensor models would have
to be matched to real hardware.

\section{Conclusion}
\label{sec:conclusion}

We have presented a quaternion error-state Kalman filter fusing up to nine aiding sensors for inertial
navigation, in the standard local-error formulation, with the error kinematics derived from first
principles, discretized with care, and each of the nine measurement Jacobians given and
finite-difference-verified. The emphasis throughout is that correctness is \emph{demonstrated}, not
asserted: the full estimation loop passes a Monte-Carlo NEES consistency gate in aggregate and over
time, rejects outliers with a companion NIS test, and reproduces the inertial-navigation and
observability behaviour theory predicts---unbounded dead-reckoning drift, and graceful recovery from any
single absolute position aid, including GPS-denied ranging. A recorded dataset, a replay tool, and
figure data emitted by the filter itself make every result reproducible, and the complete system runs
unchanged from the command-line gate to a real-time browser visualization.

\appendix
\section{The \texorpdfstring{$SO(3)$}{SO(3)} Exponential and Logarithm}
\label{app:so3}

For a rotation vector $\bm{\phi}\in\real^3$ with $\theta=\lVert\bm{\phi}\rVert$ and
$\hat{\bm{\phi}}=\bm{\phi}/\theta$, the group exponential is Rodrigues' formula
\begin{equation}
  \Exp(\bm{\phi}) = \exp(\skewm{\bm{\phi}}) = \I + \frac{\sin\theta}{\theta}\skewm{\bm{\phi}}
  + \frac{1-\cos\theta}{\theta^2}\skewm{\bm{\phi}}^2,
\end{equation}
with the small-angle limit $\I+\skewm{\bm{\phi}}$ (used with a series near $\theta=0$ to avoid the
removable singularity); the corresponding unit quaternion is
$\Exp(\tfrac12\bm{\phi}) = \big(\cos\tfrac{\theta}{2},\ \sin\tfrac{\theta}{2}\,\hat{\bm{\phi}}\big)$.
The orientation block of \eqref{eq:F} is $\Exp(-\bomega\,\Delta t)$ evaluated with this formula rather
than truncated. The inverse (the logarithmic map) recovers the rotation vector from a unit quaternion
$(q_w,\bm{q}_v)$ as
\begin{equation}
  \Log(q_w,\bm q_v) = 2\,\atantwo\!\big(\lVert\bm{q}_v\rVert,\ q_w\big)\,
  \frac{\bm{q}_v}{\lVert\bm{q}_v\rVert},
\end{equation}
and defines the $\boxminus$ used to form the orientation component of the estimation error in
Section~\ref{sec:nees}. Both maps are unit-tested for round-trip accuracy and against known rotations.

\section{Quaternion Algebra and Conventions}
\label{app:quat}

We use Hamilton quaternions with the real part first, $\bq=(q_w,\bm q_v)$, and the product
\begin{equation}
  \bq_1\otimes\bq_2 =
  \begin{pmatrix}
    q_{1w}q_{2w} - \bm q_{1v}\!\cdot\!\bm q_{2v} \\[2pt]
    q_{1w}\bm q_{2v} + q_{2w}\bm q_{1v} + \bm q_{1v}\times\bm q_{2v}
  \end{pmatrix},
\end{equation}
which is right-handed ($\bm i\bm j=\bm k$). The rotation matrix mapping body to world is
\begin{equation}
  \R(\bq) = (q_w^2 - \bm q_v^{\!\top}\bm q_v)\I + 2\,\bm q_v\bm q_v^{\!\top}
  + 2\,q_w\skewm{\bm q_v},
\end{equation}
and satisfies $\R(\bq_1\otimes\bq_2)=\R(\bq_1)\R(\bq_2)$. The local (body-frame) perturbation
$\bq\otimes\Exp(\tfrac12\dtheta)$ of \eqref{eq:localerror} corresponds to a right multiplication of
$\R$, $\R\,\Exp(\dtheta)$, which is why every orientation-dependent measurement Jacobian in
Section~\ref{sec:meas} enters through $\R^\top$ and the pattern of \eqref{eq:rotjac}. The JPL convention
of~\cite{trawny2005} instead orders the product left-handedly and transposes $\R(\bq)$; the two
conventions describe the same physics but differ by transposes that, if mixed, produce sign errors in
exactly the coupling terms of \eqref{eq:Acont}. Fixing one convention---here, Hamilton---and testing the
rotation composition against direction-cosine matrices is the practical guard against that class of bug.

\begin{thebibliography}{99}

\bibitem{kalman1960}
R.~E. Kalman, ``A new approach to linear filtering and prediction problems,''
\emph{J. Basic Eng.}, vol.~82, no.~1, pp.~35--45, 1960.

\bibitem{sola2017}
J.~Sol\`a, ``Quaternion kinematics for the error-state Kalman filter,'' arXiv:1711.02508, 2017.

\bibitem{trawny2005}
N.~Trawny and S.~I. Roumeliotis, ``Indirect Kalman filter for 3D attitude estimation,''
Univ.\ of Minnesota, Tech.\ Rep.\ 2005-002, 2005.

\bibitem{lefferts1982}
E.~J. Lefferts, F.~L. Markley, and M.~D. Shuster, ``Kalman filtering for spacecraft attitude
estimation,'' \emph{J. Guidance, Control, and Dynamics}, vol.~5, no.~5, pp.~417--429, 1982.

\bibitem{markley2003}
F.~L. Markley, ``Attitude error representations for Kalman filtering,''
\emph{J. Guidance, Control, and Dynamics}, vol.~26, no.~2, pp.~311--317, 2003.

\bibitem{crassidis2007}
J.~L. Crassidis, F.~L. Markley, and Y.~Cheng, ``Survey of nonlinear attitude estimation methods,''
\emph{J. Guidance, Control, and Dynamics}, vol.~30, no.~1, pp.~12--28, 2007.

\bibitem{shuster1993}
M.~D. Shuster, ``A survey of attitude representations,''
\emph{J. Astronautical Sciences}, vol.~41, no.~4, pp.~439--517, 1993.

\bibitem{barshalom2001}
Y.~Bar-Shalom, X.~R. Li, and T.~Kirubarajan, \emph{Estimation with Applications to Tracking and
Navigation}. New York: Wiley, 2001.

\bibitem{maybeck1979}
P.~S. Maybeck, \emph{Stochastic Models, Estimation, and Control, Vol.~1}. New York: Academic Press, 1979.

\bibitem{gelb1974}
A.~Gelb (ed.), \emph{Applied Optimal Estimation}. Cambridge, MA: MIT Press, 1974.

\bibitem{vanloan1978}
C.~F. Van Loan, ``Computing integrals involving the matrix exponential,''
\emph{IEEE Trans. Automatic Control}, vol.~23, no.~3, pp.~395--404, 1978.

\bibitem{farrell2008}
J.~A. Farrell, \emph{Aided Navigation: GPS with High Rate Sensors}. New York: McGraw-Hill, 2008.

\bibitem{groves2013}
P.~D. Groves, \emph{Principles of GNSS, Inertial, and Multisensor Integrated Navigation Systems},
2nd~ed. Boston: Artech House, 2013.

\bibitem{titterton2004}
D.~H. Titterton and J.~L. Weston, \emph{Strapdown Inertial Navigation Technology}, 2nd~ed. IET, 2004.

\bibitem{kok2017}
M.~Kok, J.~D. Hol, and T.~B. Sch\"on, ``Using inertial sensors for position and orientation
estimation,'' \emph{Foundations and Trends in Signal Processing}, vol.~11, no.~1--2, pp.~1--153, 2017.

\bibitem{mourikis2007}
A.~I. Mourikis and S.~I. Roumeliotis, ``A multi-state constraint Kalman filter for vision-aided
inertial navigation,'' in \emph{IEEE ICRA}, 2007, pp.~3565--3572.

\bibitem{qin2018}
T.~Qin, P.~Li, and S.~Shen, ``VINS-Mono: A robust and versatile monocular visual-inertial state
estimator,'' \emph{IEEE Trans. Robotics}, vol.~34, no.~4, pp.~1004--1020, 2018.

\bibitem{barrau2017}
A.~Barrau and S.~Bonnabel, ``The invariant extended Kalman filter as a stable observer,''
\emph{IEEE Trans. Automatic Control}, vol.~62, no.~4, pp.~1797--1812, 2017.

\bibitem{kaess2012}
M.~Kaess, H.~Johannsson, R.~Roberts, V.~Ila, J.~J. Leonard, and F.~Dellaert, ``iSAM2: Incremental
smoothing and mapping using the Bayes tree,'' \emph{Int. J. Robotics Research}, vol.~31, no.~2,
pp.~216--235, 2012.

\bibitem{forster2017}
C.~Forster, L.~Carlone, F.~Dellaert, and D.~Scaramuzza, ``On-manifold preintegration for real-time
visual--inertial odometry,'' \emph{IEEE Trans. Robotics}, vol.~33, no.~1, pp.~1--21, 2017.

\bibitem{huang2010}
G.~P. Huang, A.~I. Mourikis, and S.~I. Roumeliotis, ``Observability-based rules for designing
consistent EKF SLAM estimators,'' \emph{Int. J. Robotics Research}, vol.~29, no.~5, pp.~502--528, 2010.

\bibitem{hesch2014}
J.~A. Hesch, D.~G. Kottas, S.~L. Bowman, and S.~I. Roumeliotis, ``Consistency analysis and improvement
of vision-aided inertial navigation,'' \emph{IEEE Trans. Robotics}, vol.~30, no.~1, pp.~158--176, 2014.

\bibitem{martinelli2012}
A.~Martinelli, ``Vision and IMU data fusion: Closed-form solutions for attitude, speed, absolute scale,
and bias determination,'' \emph{IEEE Trans. Robotics}, vol.~28, no.~1, pp.~44--60, 2012.

\bibitem{julier2004}
S.~J. Julier and J.~K. Uhlmann, ``Unscented filtering and nonlinear estimation,''
\emph{Proc. IEEE}, vol.~92, no.~3, pp.~401--422, 2004.

\bibitem{mueller2015}
M.~W. Mueller, M.~Hamer, and R.~D'Andrea, ``Fusing ultra-wideband range measurements with
accelerometers and rate gyroscopes for quadrocopter state estimation,'' in \emph{IEEE ICRA}, 2015,
pp.~1730--1736.

\bibitem{honegger2013}
D.~Honegger, L.~Meier, P.~Tanskanen, and M.~Pollefeys, ``An open source and open hardware embedded
metric optical flow CMOS camera for indoor and outdoor applications,'' in \emph{IEEE ICRA}, 2013,
pp.~1736--1741.

\bibitem{burri2016}
M.~Burri et~al., ``The EuRoC micro aerial vehicle datasets,'' \emph{Int. J. Robotics Research},
vol.~35, no.~10, pp.~1157--1163, 2016.

\bibitem{schubert2018}
D.~Schubert, T.~Goll, N.~Demmel, V.~Usenko, J.~St\"uckler, and D.~Cremers, ``The TUM VI benchmark for
evaluating visual-inertial odometry,'' in \emph{IEEE/RSJ IROS}, 2018, pp.~1680--1687.

\bibitem{delmerico2019}
J.~Delmerico, T.~Cieslewski, H.~Rebecq, M.~Faessler, and D.~Scaramuzza, ``Are we ready for autonomous
drone racing? The UZH-FPV drone racing dataset,'' in \emph{IEEE ICRA}, 2019, pp.~6713--6719.

\bibitem{geiger2013}
A.~Geiger, P.~Lenz, C.~Stiller, and R.~Urtasun, ``Vision meets robotics: The KITTI dataset,''
\emph{Int. J. Robotics Research}, vol.~32, no.~11, pp.~1231--1237, 2013.

\end{thebibliography}

\end{document}
